Climate variability indices for ecological and crop models in R: the `climatrends` package

### 2020-10-02 inst/paper/paper.Rmd
paper

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Magne Neby
    affiliation: "3"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
  - name: Department of Forest and Field Sciences, Inland Norway University of Applied Sciences, 2480 Koppang, Norway
    index: 3
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: false
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

```{r table2, echo=FALSE, results="asis"}
index <- as.data.frame(matrix(
  c("maxDT", "Maximun day temperature", "°C",
    "minDT", "Minimum day temperature", "°C",
    "maxNT", "Maximun night temperature", "°C",
    "minNT", "Minimum night temperature", "°C",
    "DTR", "Diurnal temperature range (mean difference between DT and NT)", "°C",
    "SU", "Summer days, number of days with maximum temperature > 30 °C", "days", 
    "TR", "Tropical nights, number of nights with maximum temperature > 25 °C", "days",
    "CFD", "Consecutive frosty days, number of days with temperature < 0 °C", "days",
    "WSDI","Maximum warm spell duration, consecutive days with temperature > 90th percentile", "days",
    "CSDI","Maximum cold spell duration, consecutive nights with temperature < 10th percentile","days",
    "T10p","The 10th percentile of night tempeture","°C",
    "T90p","The 90th percentile of day tempeture","°C",
    "MLDS", "Maximum length of consecutive dry day, rain < 1 mm", "days",
    "MLWS", "Maximum length of consecutive wet day, rain >= 1 mm", "days",
    "R10mm", "Heavy precipitation days 10 >= rain < 20 mm", "days",
    "R20mm", "Very heavy precipitation days rain >= 20", "days",
    "Rx1day", "Maximum 1-day precipitation", "mm",
    "Rx5day", "Maximum 5-day precipitation", "mm",
    "R95p", "Total precipitation when rain > 95th percentile", "mm",
    "R99p", "Total precipitation when rain > 99th percentile", "mm",
    "Rtotal", "Total precipitation in wet days, rain >= 1 mm", "mm",
    "SDII", "Simple daily intensity index, total precipitation divided by the number of wet days", "mm/days"),
  nrow = 22, ncol = 3, byrow = TRUE
))

@@

The degree-days spanned from 54 to 100 days as shown in Fig. 2a. For simplicity we take the average per season and use this vector to compute the temperature indices.

@@

Across-season distribution of maxNT captured for each sample plot in this experiment is shown in Fig. 2b. The data has a bimodal distribution which is reflected in the splitting value (18.7 $^\circ C$) for the Plackett-Luce trees in Fig. 2c. The upper node splits with 49 summer days (SU). We can interpret these results as that differences in growing performance of common beans is led by a considerable amount of diurnal temperature above a warmer threshold of 30 $^\circ C$ (in this case >70% of the growing days) and warmer nights (> 18.7 $^\circ C$).

```{r fig_cbean, fig.cap="Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in three different nodes splitted with the summer days (day temperature > 30 °C) and maximum night temperature (°C). Note: the first season (primera, Pr) spans from May to August, the second (postrera, Po) from September to October, and the third (apante, Ap) from November to January.", out.width = '80%', echo=FALSE}
knitr::include_graphics("cbean.png")
```

@@

Pick some random points in Norway and Sweden and check how the trends on temperature indices over the last 20 years.

## Impact of early thaw on vole populations

From [Johnsen et al (2016)](https://doi.org/10.1002/ece3.2635) and [Boonstra et al (2016)](http://bioscience.oxfordjournals.org/content/early/2016/07/19/biosci.biw080.full.pdf). The snowpack of European boreal forest contains relatively dense snow because of snow falling at temperatures just below freezing and because of the formation of ice crusts owing to thaw–freeze cycles during the relatively mild winters. The formation of ice due to freeze–thaw cycles during winter may be limiting vole availability to bilberries and other edible food items by encrusting them in ice and therefore limiting the peak phase of the cycles. If there falls snow in autumn but then it becomes above zero and the snow starts melting but then it starts freezing again, an ice layer is formed which has important implications in ecology. This might be calculated / predicted with the `late_frost()` function, but then applied as 'early thaw' function.

@@

Integration with other datasets as they become available in `R` via API client packages. New indices related to the physiology of crops to be implemented while I work on the rice data.

@@

# Authors' contributions

KdS and JvE conceived the ideas of implementing an `R` package. JvE and SØS assisted in the implementation of methods. KdS wrote and documented the software. [add other contributions...]

@@

The source code has been archived at [**add link**] as `climatrends` version **X.Y.Z**. To explore the latest functionalities of `climatrends`, please check the package's updates at CRAN (https://cran.r-project.org/package=climatrends).

>>>>

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Magne Neby
    affiliation: "3"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
  - name: Department of Forest and Field Sciences, Inland Norway University of Applied Sciences, 2480 Koppang, Norway
    index: 3
citation_author: de Sousa et. al.
date: "A preprint"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: false
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

```{r table2, echo=FALSE, results="asis"}
index <- as.data.frame(matrix(
  c("maxDT", "Maximun day temperature", "°C",
    "minDT", "Minimum day temperature", "°C",
    "maxNT", "Maximun night temperature", "°C",
    "minNT", "Minimum night temperature", "°C",
    "DTR", "Diurnal temperature range (mean difference between DT and NT)", "°C",
    "SU", "Summer days, number of days with maximum temperature > 30 °C", "days", 
    "TR", "Tropical nights, number of nights with maximum temperature > 25 °C", "days",
    "CFD", "Consecutive frosty days, number of days with temperature < 0 °C", "days",
    "WSDI","Maximum warm spell duration, consecutive days with temperature > 90th percentile", "days",
    "CSDI","Maximum cold spell duration, consecutive nights with temperature < 10th percentile","days",
    "T10p","The 10th percentile of night temperature","°C",
    "T90p","The 90th percentile of day temperature","°C",
    "MLDS", "Maximum length of consecutive dry day, rain < 1 mm", "days",
    "MLWS", "Maximum length of consecutive wet day, rain >= 1 mm", "days",
    "R10mm", "Heavy precipitation days 10 >= rain < 20 mm", "days",
    "R20mm", "Very heavy precipitation days rain >= 20", "days",
    "Rx1day", "Maximum 1-day precipitation", "mm",
    "Rx5day", "Maximum 5-day precipitation", "mm",
    "R95p", "Total precipitation when rain > 95th percentile", "mm",
    "R99p", "Total precipitation when rain > 99th percentile", "mm",
    "Rtotal", "Total precipitation in wet days, rain >= 1 mm", "mm",
    "SDII", "Simple daily intensity index, total precipitation divided by the number of wet days", "mm/days"),
  nrow = 22, ncol = 3, byrow = TRUE
))

@@

The degree-days spanned from 54 to 100 days as shown in Fig. 1a. For simplicity we take the average per season and use this vector to compute the temperature indices.

@@

Across-season distribution of maxNT captured for each sample plot in this experiment is shown in Fig. 1b. The data has a bimodal distribution which is reflected in the splitting value (18.7 $^\circ C$) for the Plackett-Luce trees in Fig. 1c. The upper node splits with 49 summer days (SU). We can interpret these results as that differences in growing performance of common beans is led by a considerable amount of diurnal temperature above a warmer threshold of 30 $^\circ C$ (in this case >70% of the growing days) and warmer nights (> 18.7 $^\circ C$).

```{r fig_cbean, fig.cap="Fig. 1. Application of climatrends functions to support the analysis of a citizen-science data testing 11 common bean varieties in Nicaragua. (A) Days required to reach 900 growing-degree days from planting date calculated using the function GDD(). (B) Maximum night temperature (°C) distributed across seasons computed using the function temperature(). (C) Plackett-Luce Tree showing the probability of one common bean variety has to win against the others (axys X) in three different nodes splitted with the summer days (day temperature > 30 °C) and maximum night temperature (°C). Note: the first season (primera, Pr) spans from May to August, the second (postrera, Po) from September to October, and the third (apante, Ap) from November to January.", out.width = '80%', echo=FALSE, fig.align="center"}
knitr::include_graphics("cbean.png")
```

@@

We randomly selected 100 points in hexagonal within the coordinates 7$^\circ$ and 17$^\circ$ W, and 59 $^\circ$ and 63 $^\circ$ N, that comprises Norway and Sweden before the Arctic Circle. We compute the temperature indices from 2000-01-01 to 2019-12-31 using the function `temperature()` with the method for objects of class 'sf'. The temperature data is fetched from from the NASA Langley Research Center POWER Project funded through the NASA Earth Science Directorate Applied Science Program (<https://power.larc.nasa.gov/>), using the R package `nasapower` [@Sparks2018].

```{r nordic, message=FALSE, echo=TRUE, eval=FALSE}
library("climatrends")
library("sf")
library("nasapower")

# create a polygon within the coordinates 7, 17, 59, 63
e <- matrix(c(7, 59, 17, 59, 17, 63,
              7, 63, 7, 59), 
            nrow = 5, ncol = 2, byrow = TRUE)

e <- st_polygon(list(e))

# sample 100 points in the hexagonal type
p <- st_sample(e, 100, type = "hexagonal")
p <- st_as_sf(p, crs = 4326)

# compute the temperature indices using the random points 
temp <- temperature(p, 
                    day.one = "2000-01-01", 
                    last.day = "2019-12-31", 
                    timeseries = TRUE, 
                    intervals = 365)

We then select the indices CSDI (cold spell duration of night temperature), WSDI (warm spell duration of day temperature), and their associated indices the T10p (the 10th percentile of night temperature) and T90p (the 90th percentile of day temperature), in Figure 2. Plots are generated with `ggplot2` [@ggplot2] and `patchwork` [@patchwork].

The trends show a decrease in the cold spell duration (number of consecutive cold nights bellow the 10th percentile) and warm spell duration (number of consecutive warm days above the 90th percentile). However, the values of the percentiles show an increase over the time series. The T10p index shows a decrease around the year of 2010, but again rises up to the a value around the -10 $^\circ$C, meaning that the could nights are becoming a bit warmer over the time. The T90p index also shows an increase in the temperature across the sampled area, with the average 90th percentile rising from ~ 16 $^\circ$C to ~ 18 $^\circ$C over the time series.

```{r fig_nordic, fig.cap="Fig. 2. Trends in temperature indices across Southern Norway and Sweeden from 2000 to 2019. CSDI, maximum cold spell duration, consecutive nights with temperature < 10th percentile. WSDI, maximum warm spell duration, consecutive days with temperature > 90th percentile. T10p, the 10th percentile of night temperature. T90p, the 90th percentile of day temperature. Red line indicates the historical mean of each index in the time series. Blue line indicates the smoothed trends in each index using the 'loess' method.", out.width = '80%', echo=FALSE, fig.align="center"}
knitr::include_graphics("nordic.png")
```

@@

The package can support the integration with other datasets as they become available in `R` via API client packages. Also new indices related to the physiology of crops could be implemented.

@@

@@

To explore the latest functionalities of `climatrends`, please check the package's updates at CRAN (https://cran.r-project.org/package=climatrends).
